Observing the United States' Homelessness Crisis through Mathematical Methods

### 2024-03-04 paper.tex
done

\section*{Executive Summary}
\blindtext

@@

\subsection{Restatement of the Problem}
The first problem asks us to develop a mathematical model to predict changes to the housing supply over the next 50
years in two cities of our choosing: Seattle, Washington; and Albequerque, New Mexico.

@@

\noindent
\textit{Differences in median housing costs and median income per capita across regions in both Albequerque and Seattle
will not be considered in determining housing availability.} \\

@@

\subsection{Model Development}
For our housing availability model, we choose to use a simple vector machine that regresses historical housing availiability
data for both Albequerque and Seattle, and from there, we cross-reference this data with historical economic trends
across the United States to best model housing availability over time given cyclical economic trends.

\noindent
To begin, we trained the model on the following datasets for housing availability in Albequerque and Seattle and the
national GDP of the United States.

\begin{table}[H]
  \centering
  \begin{tabular}{|c c c c|} 
    \hline 
    Year & Total housing units & Occupied units & Vacant units \\ [0.5ex]
    \hline
    2010  &  234,891 & 217,256 & 17,635 \\
    2011  &  237,735 & 220,060 & 17,675 \\
    2012  &  239,718 & 222,584 & 17,134 \\
    2013  &  240,277 & 222,491 & 17,786 \\
    2014  &  240,961 & 222,868 & 18,093 \\
    2015  &  241,326 & 222,098 & 19,228 \\
    2016  &  242,070 & 221,320 & 20,750 \\
    2017  &  243,402 & 221,119 & 22,283 \\
    2018  &  244,382 & 222,748 & 21,634 \\
    2019  &  245,476 & 224,166 & 21,310 \\
    2020  &  247,926 & 229,701 & 18,225 \\
    2021  &  252,924 & 236,191 & 16,733 \\
    2022  &  255,178 & 239,800 & 15,378 \\ [1ex] 
    \hline
  \end{tabular}
  \caption{Housing Statistics in Albequerque, New Mexico \cite{Census2010ACSDP1Y2010.DP04}.}
\end{table}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\textwidth]{vacancy}
  \caption{The number of vacancies in Albequerque and Seattle from 2010 to 2022 \cite{Census2010ACSDP1Y2010.DP04,Census2010ACSDP5Y2010.DP04}.}
\end{figure}

@@

\subsection{Restatement of the Problem}
The first problem asks us to develop a mathematical model to predict changes to the homeless population over the next 50
years in two cities of our choosing: Seattle, Washington; and Albequerque, New Mexico.

@@

\begin{table}[H]
  \centering
  \begin{tabular}{|c c c c|} 
    \hline 
    Year & Total housing units & Occupied units & Vacant units \\ [0.5ex]
    \hline
    2010 & 234,891 & 217,256 & 17,635 \\
    2011 & 237,735 & 220,060 & 17,675 \\
    2012 & 239,718 & 222,584 & 17,134 \\
    2013 & 240,277 & 222,491 & 17,786 \\
    2014 & 240,961 & 222,868 & 18,093 \\
    2015 & 241,326 & 222,098 & 19,228 \\
    2016 & 242,070 & 221,320 & 20,750 \\
    2017 & 243,402 & 221,119 & 22,283 \\
    2018 & 244,382 & 222,748 & 21,634 \\
    2019 & 245,476 & 224,166 & 21,310 \\
    2020 & 247,926 & 229,701 & 18,225 \\
    2021 & 252,924 & 236,191 & 16,733 \\
    2022 & 255,178 & 239,800 & 15,378 \\ [1ex] 
    \hline
  \end{tabular}
  \caption{Housing Statistics in Albequerque, New Mexico \cite{Census2010ACSDP1Y2010.DP04}.}
\end{table}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{homeless-vs-year-alb}
  \caption{Total Homeless by Year in Albequerque, New Mexico}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{homeless-prop-vs-year-alb}
  \caption{Percentage of the Population that is Homeless in Albequerque, New Mexico}
\end{figure}

\noindent
Therefore, for Albequerque, let $A_{T}(x)$ be total number of homeless people in Albequerque, where $x$ is the current year (which is absolute from 0, such as 2024).

@@

\noindent
Additionally, let $A_{\%}(x)$ be the percentage of Albequerque's population that is homeless, where $x$ is the current year (which is absolute from 0, such as 2024).

@@

@@

\subsubsection{Albequerque, New Mexico}

@@

\subsection{Restatement of the Problem}
\subsection{Assumptions and Justifications}
\subsection{Model Development}
\subsection{Results}
\subsection{Reflecting on the Model}

>>>>

\section*{Executive Summary}
For the first part of our paper, we were asked to develop a mathematical model to predict changes to the housing
supply over the next 50 years in two cities of our choosing, of which we chose Seattle, Washington and Albuquerque,
New Mexico. To solve this, we developed a linear model through regressing historical housing supply data across both
cities, and determined that it proves weak at accounting for large unforseen events like war, epidemics, or natural
disasters while still successfully guiding public policy decisions under the assumption that such instances rarely, if
ever, occur. \\

\noindent
For the second part of our paper, we are asked to develop a similar model to track the rate of homelessness across
both cities. Again, we regressed historical homelessness data to develop a linear model, except in these models, we
chose to develop models designed for each city rather than one aggregate model due to the completely different
homelessness situations across each city. \\

\noindent
Lastly, for the third and final part of our paper, we are asked to develop a model to best aid local governments in
developing public policy to combat homelessness. We conclude that, based on our results from part 2, public policy
decisions made by Albuquerque will be the most successful, and as such, we chose to mathematically model the effects
of their public policy on homelessness populations. From there, we reflected on our model and determined that it, yet
again, fails to account for difficult-to-foresee situations but serves well as a general guideline for local
governments when developing public policy decisions.

@@

\subsection{Restatement of the Problem}
The first problem asks us to develop a mathematical model to predict changes to the housing supply over the next 50
years in two cities of our choosing: Seattle, Washington; and Albuquerque, New Mexico.

@@

\noindent
\textit{Differences in median housing costs and median income per capita across regions in both Albuquerque and Seattle
will not be considered in determining housing availability.} \\

@@

\subsection{Model Development}
For our housing availability model, we choose to use a simple vector machine that regresses historical housing availiability
data for both Albuquerque and Seattle, and from there, we cross-reference this data with historical economic trends
across the United States to best model housing availability over time given cyclical economic trends.

\noindent
To begin, we trained the model on the following datasets for housing availability in Albuquerque and Seattle and the
national GDP of the United States.

\begin{table}[H]
  \centering
  \begin{tabular}{|c c c c|} 
    \hline 
    Year & Total housing units & Occupied units & Vacant units \\ [0.5ex]
    \hline
    2010  &  234,891 & 217,256 & 17,635 \\
    2011  &  237,735 & 220,060 & 17,675 \\
    2012  &  239,718 & 222,584 & 17,134 \\
    2013  &  240,277 & 222,491 & 17,786 \\
    2014  &  240,961 & 222,868 & 18,093 \\
    2015  &  241,326 & 222,098 & 19,228 \\
    2016  &  242,070 & 221,320 & 20,750 \\
    2017  &  243,402 & 221,119 & 22,283 \\
    2018  &  244,382 & 222,748 & 21,634 \\
    2019  &  245,476 & 224,166 & 21,310 \\
    2020  &  247,926 & 229,701 & 18,225 \\
    2021  &  252,924 & 236,191 & 16,733 \\
    2022  &  255,178 & 239,800 & 15,378 \\ [1ex] 
    \hline
  \end{tabular}
  \caption{Housing Statistics in Albuquerque, New Mexico \cite{Census2010ACSDP1Y2010.DP04}.}
\end{table}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.9\textwidth]{vacancy}
  \caption{The number of vacancies in Albuquerque and Seattle from 2010 to 2022 \cite{Census2010ACSDP1Y2010.DP04,Census2010ACSDP5Y2010.DP04}.}
\end{figure}

@@

\subsection{Restatement of the Problem}
The first problem asks us to develop a mathematical model to predict changes to the homeless population over the next 50
years in two cities of our choosing: Seattle, Washington; and Albuquerque, New Mexico.

@@

\begin{table}[H]
  \centering
  \begin{tabular}{|c c c|} 
    \hline 
    Year & Total Population & Total Homeless \\ [0.5ex]
    \hline
    2007 & -       & 1,276 \\
    2008 & -       & 1,276 \\
    2009 & -       & 2,002 \\
    2010 & 531,403 & 2,002 \\
    2011 & 539,000 & 1,639 \\
    2012 & 545,083 & 1,431 \\
    2013 & 549,812 & 1,171 \\
    2014 & 553,576 & 1,254 \\
    2015 & 556,092 & 1,287 \\
    2016 & 556,859 & 1,222 \\
    2017 & 556,718 & 1,318 \\
    2018 & 559,202 & 1,340 \\
    2019 & 559,374 & 1,524 \\
    2020 & 560,447 & 1,586 \\
    2021 & 562,336 & 1,567 \\
    2022 & 562,551 & 1,277 \\
    \hline
  \end{tabular}
  \caption{Total Population by Year in Albuquerque \cite{Census2022ACSST1Y2022.S0101,HOMELESS-STATS}.}
\end{table}

\begin{table}[H]
  \centering
  \begin{tabular}{|c c|} 
    \hline 
    Year & Total Homeless \\ [0.5ex]
    \hline

    \hline
  \end{tabular}
  \caption{}
\end{table}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{homeless-vs-year-alb}
  \caption{Total Homeless by Year in Albuquerque, New Mexico}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{homeless-prop-vs-year-alb}
  \caption{Percentage of the Population that is Homeless in Albuquerque, New Mexico}
\end{figure}

\noindent
Therefore, for Albuquerque, let $A_{T}(x)$ be total number of homeless people in Albequerque, where $x$ is the current year (which is absolute from 0, such as 2024).

@@

\noindent
Additionally, let $A_{\%}(x)$ be the percentage of Albuquerque's population that is homeless, where $x$ is the current year (which is absolute from 0, such as 2024).

@@

\begin{table}[H]
  \centering
  \begin{tabular}{|c c c|} 
    \hline 
    Year & Total Population & Total Homeless \\ [0.5ex]
    \hline
    2007 & -       & 7,902 \\
    2008 & -       & 8,501 \\
    2009 & -       & 8,952 \\
    2010 & 595,240 & 9,022 \\
    2011 & 603,174 & 8,972 \\
    2012 & 612,916 & 8,899 \\
    2013 & 624,681 & 9,106 \\
    2014 & 637,850 & 8,949 \\
    2015 & 653,017 & 10,122 \\
    2016 & 668,849 & 10,730 \\
    2017 & 688,245 & 11,643 \\
    2018 & 708,823 & 12,112 \\
    2019 & 724,305 & 11,199 \\
    2020 & 741,251 & 11,751 \\
    2021 & 726,054 & 5,183 \\
    2022 & 734,603 & 13,368 \\
    \hline
  \end{tabular}
  \caption{Total Population by Year in Seattle \cite{Census2022ACSST1Y2022.S0101,HOMELESS-STATS}.}
\end{table}

@@

\subsubsection{Albuquerque, New Mexico}

@@

\subsection{Restatement of the Problem}
The third and final question asks us to utilize the results from at least one of our previous models to first develop a final
model that can be used to best aid a city in combatting the homelessness crisis then analyze the model for its
strengths and weaknesses given situations like natural disasters, economic recessions, or migrant populations.

\subsection{Assumptions and Justifications}
\textit{Job availability will be stimulated primarily through public policy from the government.} \\

\noindent
The intent of this model is to understand how public policy ought to reflect into homelessness statistics without
heavy consideration to external factors, such as the growth or shrinking of privately funded opportunities
supplied for homeless people (such as charity or jobs). Such consideration would require more narrow focus on cyclical
economic trends, when in reference to the success of implementing simple public policy, such methods would be rendered
unnecessary. For the sake of brevity, lawmakers merely need to understand the general impacts of their legislation rather than their impact in
context of the entire economic situation of their region.

\subsection{Model Development}

We developed our model based on the predictions outlined in recent public policy decisions made by Albuquerque city
officials of combatting homelessness through policy-funded stimulation of job opportunities directly for homeless
shelter residents\cite{ALB-JOBS}. Given the immense success Albuquerque has had thus far in reducing homelessness year
after year (as shown in our previous results), we can conclude that their success in providing conncting opportunities
to the homeless ought to serve as the benchmark for determining public policy for addressing homeless populations
throughout the nation. \\

\noindent
Thus, we developed the following model through our previous linear regression processes with data from Albuquerque's
labor statistics \cite{ALB-LABOR}. Let $N(x)$ equal our homelessness population in Albuquerque for a given year, where $x$ is the
number of available jobs for the given market.

$$N(x) = 1,368-0.037(x)$$

This model provides a strong general basis for understanding how the availability of jobs for homeless populations
within Albequerque serves as an inverse relationship, directly combatting the homelessness crisis in an incredibly
effective manner. However, we must acknowledge that it is unreliable in consideration to unforseen events that could
divert resources from providing economic opportunities like jobs for the homeless, such as war or natural disasters,
due to lacking nuance for accounting for such occurances. Additionally, as a linear model, it will be far less
accurate for accounting for cyclical movements in the economy, acting as more of an aggregate measure for combatting
homelessness through job opportunities than a guaranteed predictor of homelessness statistics.
